A Quasi-Genuine FIFO Total Order Multicast Primitive

### 2011-06-09 trunk/textos/optpaxos/draftv3.tex
shit's getting complete

A more formal description of the protocol is given in Algorithm \ref{algorithm:quasi}. We consider that three primitives are given: $getTime()$, which returns the current value of the local wallclock; $Propose(k, val)$, which proposes a value $val$ for the consensus instance of id $k$ within its group; and also $Decide(k, val)$, which is called when the consensus instance of id $k$ finishes. $Decide(k, val)$ is called for all the processes of the group that initiated it, when they learn that the value $val$ has been agreed upon in instance of id $k$. For the sake of simplicity, we assume that, for consensus instances within the same group, the values are decided in the same order of the instances id's\footnote{This can be easily done by delaying the callback of $Decide(k, val)$ while there is some unfinished consensus instance of id $k': k' < k$ from the same group.}. Finally, we also use a FIFO reliable multicast primitive \rmcast{$m$}{$groupSet$}, which \rmd{}s $m$ to all the processes in all the groups in $groupSet$ in one communication step, in FIFO order (e.g. the one described in \cite{ufrmcast1delta}).

@@

\caption{\cmsend{m} -- executed by every process $p$ from group $G$}
\label{algorithm:quasi}
\end{distribalgo}
\end{algorithm}

@@

The problem with Algorithm \ref{algorithm:quasi} is that it does not guarantee liveness when a group has no message to receive from some other group and then keeps waiting for a new message to increase the barrier value and proceed with the delivery of new messages. However, liveness can be easily provided by sending periodic empty messages from $G$ to each $G' \in receiversFrom(G)$ to which no message has been sent for a specified time threshold \textit{barrierThreshold}. Algorithm \ref{algorithm:liveness} describes this. When a message $m$ from group $G$ to some other group $G'$ has been \rmc{} by $p \in G$, $p$ knows that the other processes of $G$ did the same and that $m$ will be eventually received by the processes of $G'$, serving as barrier from $G$ to $G'$ (l. \ref{algline:incbar} of Algorithm \ref{algorithm:quasi}). However, when there is a long period after the last time when such kind of message has been created, $p$ decides to create some empty message to send to the processes of $G'$ with the sole purpose of increasing their barrier values and allow for the delivery of possibly blocked messages in $G'$.

@@

\blankline
\caption{Achieving liveness by sending periodic messages; executed by every process $p$ of group $G$}
\label{algorithm:liveness}
\end{distribalgo} 
\end{algorithm}

@@

It is necessary to guarantee that a $null$ message will eventually be proposed, decided, and a barrier will be sent to some group which might be needing it, so that progression is guaranteed. Therefore, this kind of messages are created by every process in the group, since if any one of them does not have it, such message might never be decided. %Besides, they are excluded from the $propPending$ list only when decided in the group (l. \ref{algline:nullstays} of Algorithm \ref{algorithm:quasi}). Since they are obviously never delivered to the application, they can remain in such list and be decided even after some message with higher timestamp is decided. Not removing them unless they are decided ensures that they will be sent and that other groups will be able to increase their barrier values. % On the other hand, this brings another problem: if every process in $G$ creates a different $null$ message, many $null$ messages might have to be decided before a meaningful message.
To prevent the multiple $null$ messages -- created by different processes within the group -- of being decided, they could be created in a way such that the different processes can somehow figure out that two different $null$ messages are equivalent\footnote{This could be done by assuming no timestamp collisions and by using them to uniquely identify messages. Then, only one of the messages created with the same timestamp (l. \ref{algline:samenullid} of Algorithm \ref{algorithm:liveness}) would be decided.}.

However, there is a way to provide liveness withouth such a possibly long delay for the conservative delivery of messages, although that would imply creating more messages and making deeper changes in the delivery algorithm. Let $blockers(m)$ be defined as the set of groups whose barrier is needed in order for some group to deliver $m$. More formally, $blockers(m) =$ \mbox{$\{G_B : \exists G_{dst} \in m.dst \wedge G_B \in sendersTo(G_{dst})\}$}. The idea is that, once each group $G_B$ in $blockers(m)$ have sent a barrier $b > m.ts$ to all the groups belonging to \mbox{$m.dst \cap receiversFrom(G_B)$}, all possible destinations of $m$ can deliver it. This way, instead of relying on periodic messages, whenever a process $p$ in a group $G$ has a message $m$ to send, $p$ sends $m$ to every $p' \in m.dst \cup G \cup blockers(m)$. It is sent to the groups in $blockers(m)$, so that they know that there is a message which will be blocked until they send a proper barrier to unblock it.

%When the process $p'$ of some group $G'$ receives $m$, such that $m.src \neq G'$, $p'$ knows that there might be other groups depeding on the barrier of $G'$ to deliver $m$. For that reason, it will immediately create a $null$ message with a timestamp equal to $m.ts$ and with $null.dst = m.dst$ $\cap$ $receiversFrom(G')$. Then, $p'$ will insert such message in its $optPending$ queue -- to ensure that, once it goes to $propPending$, any message \mbox{$m': m'.ts < null.ts \wedge m'.src = G'$} is already there. As soon as $null.ts > now - w(p')$, the $null$ message will be proposed in $G'$ and, once $null$ is decided, each process of $G'$ will send a $\{null, \text{`cons'}\}$ message to each process in each group \mbox{$G \in m.dst \cap receiversFrom(G')$}. This way, any group which was waiting for a barrier from $G'$ to deliver $m$ will be able to do so as soon as it receives such $null$ message. The new delivery algorithm would be as described in \mbox{Algorithm {\ref{algorithm:ondemand}}}.

@@

% \label{algorithm:ondemand}

@@

% Now, assuming that $w(p)$ is the same for every process $p$, we have a worst case delivery latency for $m$ of $w(p) + 2\tconsm$, as the $null$ message is created and proposed in parallel with $m$. However, each sender process is sending $m$ not only to every destination process, but also to every process in its group and to every process belonging to some group of $blockers(m)$. This can create a fairly high amount of messages. Unfortunately, this has to be done to guarantee that such $null$ message will eventually be proposed, decided, and a barrier will be sent to some group which might be needing it. Although this kind of messages are excluded from the $propPending$ list only when decided in the group (l. \ref{algline:keepnull} of Algorithm \ref{algorithm:ondemand}), if at least one of the processes does not have it, it could never be decided. Again, like in Algorithm \ref{algorithm:liveness}, there might be many processes in a group proposing a $null$ message because of the same $m$. Although this is not wrong, it is inefficient, and a way to identify $null$ messages created for the same purpose should be devised\footnote{Again, we could use timestamps as unique identifiers and assume no timestamp collisions. But for that to work, instead of making $null.ts = m.ts$ (l. \ref{algline:nulltsmts} of Algorithm \ref{algorithm:ondemand}), we could define $null.ts$ as a little bit greater than $m.ts$.}.

>>>>

A more formal description of the protocol is given in Algorithm \ref{algorithm:deliveryminimal}. We consider that three primitives are given: $getTime()$, which returns the current value of the local wallclock; $Propose(k, val)$, which proposes a value $val$ for the consensus instance of id $k$ within its group; and also $Decide(k, val)$, which is called when the consensus instance of id $k$ finishes. $Decide(k, val)$ is called for all the processes of the group that initiated it, when they learn that the value $val$ has been agreed upon in instance of id $k$. For the sake of simplicity, we assume that, for consensus instances within the same group, the values are decided in the same order of the instances id's\footnote{This can be easily done by delaying the callback of $Decide(k, val)$ while there is some unfinished consensus instance of id $k': k' < k$ from the same group.}. Finally, we also use a FIFO reliable multicast primitive \rmcast{$m$}{$groupSet$}, which \rmd{}s $m$ to all the processes in all the groups in $groupSet$ in one communication step, in FIFO order (e.g. the one described in \cite{ufrmcast1delta}).

@@

\caption{\cmsend{m} -- executed by every process $p$ from group $G$}
\label{algorithm:deliveryminimal}
\end{distribalgo}
\end{algorithm}

@@

The problem with Algorithm \ref{algorithm:deliveryminimal} is that it does not guarantee liveness when a group has no message to receive from some other group and then keeps waiting for a new message to increase the barrier value and proceed with the delivery of new messages. However, liveness can be easily provided by sending periodic empty messages from $G$ to each $G' \in receiversFrom(G)$ to which no message has been sent for a specified time period. Section \ref{sec:nullperiodic} describes such approach. In section \ref{sec:nullondemand}, another approach, based on sending empty messages when requested, is presented. The trade-off between them concerns message delivery latency against number of control messages sent.

\subsubsection{Sending empty messages periodically}
\label{sec:nullperiodic}

The simplest way to provide liveness to the delivery algorithm is by sending empty messages periodically. The time threshold \textit{barrierThreshold} defines how long a process waits for a meaningful message to be sent before sending an empty message to ensure that some other group is eventually unblocked. Algorithm \ref{algorithm:nullperiodic} describes this. When a message $m$ from group $G$ to some other group $G'$ has been \rmc{} by $p \in G$, $p$ knows that the other processes of $G$ did the same and that $m$ will be eventually received by the processes of $G'$, serving as barrier from $G$ to $G'$ (l. \ref{algline:incbar} of Algorithm \ref{algorithm:deliveryminimal}). However, when there is a long period after the last time when such kind of message has been created, $p$ decides to create some empty message to send to the processes of $G'$ with the sole purpose of increasing their barrier values and allow for the delivery of possibly blocked messages in $G'$.

@@

\blankline
\caption{Achieving liveness by sending periodic messages; executed by every process $p$ of group $G$}
\label{algorithm:nullperiodic}
\end{distribalgo} 
\end{algorithm}

@@

It is necessary to guarantee that a $null$ message will eventually be proposed, decided, and a barrier will be sent to some group which might be needing it, so that progression is guaranteed. Therefore, this kind of messages are created by every process in the group, since if any one of them does not have it, such message might never be decided. %Besides, they are excluded from the $propPending$ list only when decided in the group (l. \ref{algline:nullstays} of Algorithm \ref{algorithm:deliveryminimal}). Since they are obviously never delivered to the application, they can remain in such list and be decided even after some message with higher timestamp is decided. Not removing them unless they are decided ensures that they will be sent and that other groups will be able to increase their barrier values. % On the other hand, this brings another problem: if every process in $G$ creates a different $null$ message, many $null$ messages might have to be decided before a meaningful message.
To prevent the multiple $null$ messages -- created by different processes within the group -- of being decided, they could be created in a way such that the different processes can somehow figure out that two different $null$ messages are equivalent\footnote{This could be done by assuming no timestamp collisions and by using them to uniquely identify messages. Then, only one of the messages created with the same timestamp (l. \ref{algline:samenullid} of Algorithm \ref{algorithm:nullperiodic}) would be decided.}.

\subsubsection{Requesting empty messages}
\label{sec:nullondemand}

There is a way to provide liveness with a lower delay for delivering messages, although that would imply creating more messages and making deeper changes in the delivery algorithm. Let $blockers(m)$ be defined as the set of groups whose barrier is needed in order for some group to deliver $m$. More formally, \mbox{$blockers(m) = \{G_B \neq m.src : \exists G_{dst} \in m.dst \wedge G_B \in sendersTo(G_{dst})\}$}. The idea is that, once each group $G_B$ in $blockers(m)$ has sent\footnote{Here, by `sending', we mean that the message has been received at the destination.} a barrier $b > m.ts$ to all the groups belonging to \mbox{$m.dst \cap receiversFrom(G_B)$}, all possible destinations of $m$ can deliver it. This way, instead of relying on periodic messages, whenever a process $p$ in a group $G$ has a message $m$ to send, $p$ sends $m$ to every $p' \in m.dst \cup G \cup blockers(m)$. It is sent to the groups in $blockers(m)$, so that they know that there is a message whose delivery will be blocked until they send a proper barrier to unblock it. Finally, a group never blocks the delivery of its own messages -- hence the condition $G_B \neq m.src$ -- because the message itself can be seen as a barrier from its source group.

When the process $p'$ of some group $G'$ receives $m$, such that $G' \in blockers(m)$, $p'$ knows that there might be other groups depeding on the barrier of $G'$ to deliver $m$. For that reason, it will immediately create a $null$ message with a timestamp equal to $m.ts$ and with $null.dst = m.dst$ $\cap$ $receiversFrom(G')$. Then, $p'$ will insert such message in its $optPending$ queue -- to ensure that, once it goes to $propPending$, any message \mbox{$m': m'.ts < null.ts \wedge m'.src = G'$} is already there. As soon as $null.ts > now - w(p')$, the $null$ message will be proposed in $G'$ and, once $null$ is decided, each process of $G'$ will send a $\{null, \text{`cons'}\}$ message to each process in each group \mbox{$G \in m.dst \cap receiversFrom(G')$}. This way, any group which was waiting for a barrier from $G'$ to deliver $m$ will be able to do so as soon as it receives such $null$ message. The new delivery algorithm would be as described in \mbox{Algorithm {\ref{algorithm:nullondemand}}}.

@@

% \label{algorithm:nullondemand}

@@

% Now, assuming that $w(p)$ is the same for every process $p$, we have a worst case delivery latency for $m$ of $w(p) + 2\tconsm$, as the $null$ message is created and proposed in parallel with $m$. However, each sender process is sending $m$ not only to every destination process, but also to every process in its group and to every process belonging to some group of $blockers(m)$. This can create a fairly high amount of messages. Unfortunately, this has to be done to guarantee that such $null$ message will eventually be proposed, decided, and a barrier will be sent to some group which might be needing it. Although this kind of messages are excluded from the $propPending$ list only when decided in the group (l. \ref{algline:keepnull} of Algorithm \ref{algorithm:nullondemand}), if at least one of the processes does not have it, it could never be decided. Again, like in Algorithm \ref{algorithm:nullperiodic}, there might be many processes in a group proposing a $null$ message because of the same $m$. Although this is not wrong, it is inefficient, and a way to identify $null$ messages created for the same purpose should be devised\footnote{Again, we could use timestamps as unique identifiers and assume no timestamp collisions. But for that to work, instead of making $null.ts = m.ts$ (l. \ref{algline:nulltsmts} of Algorithm \ref{algorithm:nullondemand}), we could define $null.ts$ as a little bit greater than $m.ts$.}.
